Agile Monitoring of Evolving Web Services: For the Little Man

### 2018-04-08 introductory.tex
all iz well till section IV
Edited from top to Section II (incl.)

  Python is one of the fastest growing programming languages of the moment. Flask, a Python-based web framework is the technology used by tens of thousands of web applications. These applications, from interactive websites to service APIs, lack a technology specific service monitoring solution.

  In this paper, we present \tool, a tool that provide insight into the utilization and performance of evolving Flask-based web services. We present the ease with which the tool can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.

@@

%Every system is a distributed system nowadays \cite{cavage2013there}. Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
%\ml{I'm not completely happy with this paragraph}
{\em There is no getting around it: you are building a distributed system} argues Cavage  in a recent article written for the {\em Communications of the ACM} \cite{cavage2013there}. Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.

@@

Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications. Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:

@@

For projects on a budget (e.g. research, startups) the first and the second options are often not available due to time and financial constraints. Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}

To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution. \cite{Nier12b}

As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.

@@

This paper is an extension of [Anonymous Paper by Anonymous Auhthors] which was published as a NIER paper in [Anonymous Conference]. The current paper extends that paper in the following ways: 
\begin{itemize}
  \item Providing more details about the case study in Section \ref{sec:case} and extending the data collection period with half year
  \item Novel perspectives and a better categorization of the perspectives presented in Sections \ref{sec:utilandperf} -- \ref{sec:evolution}
  \item A detailed discussion on automated outlier detection and monitoring (Section \ref{sec:outliers})
  \item An advanced discussion integration continuous integration environments in order to triangulate the live collected data with information from testing the endpoints with a constant load and introduce a mechanism for performance prediction (Section \ref{sec:testing})
  \item A more extended discussion about lessons learned while developing the tool including measurements on the overhead incurred by the presented tool \ml{(... if we make it... )}
\end{itemize}

@@

  \Fref{fig:zeeguuarch} presents a high-level view of the case study ecosystem.  The core API is implemented with Flask and Python provides and correspondingly three types of functionality: contextual translations, personalized exercise suggestions, and article recommendations. In total, the API provides a bit under 50 endpoints, out of which around a dozen are very frequently used. The development of the core API itself is a research project. One of the authors of this paper is part of the developer team that maintains and evolves the \zee API.

@@

  At the time of writing, the ecosystem consists of a reader web application, a web-based exercises platform, and a smartwatch application, which are used by more than \activeUserCount active users on a regular basis. The users come from several Dutch highschools and a language center associated with a Dutch University. Several users are using it on their own, outside of any formal educational context. The learners use \zee to study a variety of of foreign languages including German, Dutch, French. They have also diverse native languages, including: Chinese, Dutch, and English. The system can scale to many languages by delegating the machine translation tasks to third party APIs.

\subsection*{Study characteristics}

@@

  All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API. The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}. The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em icsme}}.

@@

 Flask is built on top of WSGI -- The Web Server Gateway Interface -- a calling convention for web servers to forward requests to web applications or frameworks written in Python. A Python web application based on WSGI has to have one central callable object that implements the actual application. In Flask this is an instance of the Flask class. Each Flask application has to create an instance of this class itself and pass it the name of the module. The following code snippet presents a simple ``Hello World'' web service written in Python with the help of the Flask framework.

@@

>>>>

  Python is one of the fastest growing programming languages at the moment. Flask, a Python-based web framework is the technology used by tens of thousands of web applications. These applications, from interactive websites to service APIs, lack a technology-specific service monitoring solution.

  In this paper, we present \tool, a tool that provides insight into the utilization and performance of evolving Flask-based web services. We present the ease with which the tool can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides, and point to some future challenges for similar libraries.

@@

%Every system is a distributed system nowadays \cite{cavage2013there}. Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
%\ml{I'm not completely happy with this paragraph}
{\em There is no getting around it: you are building a distributed system} argues Mark Cavage  in a recent article written for the {\em Communications of the ACM} \cite{cavage2013there}. Indeed, even the simplest second-year student project these days is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.

@@

Despite their popularity however, and to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications. Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:

@@

For projects on a budget (e.g. research, startups) the first and the second options are often not available due to time and financial constraints. Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application is missing because such solutions have no notion of versioning and no integration with the development life cycle~\cite{papazoglou2011managing}.

To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution~\cite{Nier12b}.

As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations, i.e. it was being operated without any monitoring solution.

@@

This paper is an extension of [Anonymous Paper by Anonymous Auhthors] which was published as a NIER paper in [Anonymous Conference]. The current paper extends that paper in the following ways: 
\begin{itemize}
  \item Providing more details about the case study in Section \ref{sec:case} and extending the data collection period by half a year.
  \item New perspectives and a better categorization of the perspectives are presented in Sections \ref{sec:utilandperf} -- \ref{sec:evolution}.
  \item A detailed discussion on automated outlier detection and monitoring is provided (Section \ref{sec:outliers}).
  \item An advanced discussion on the integration with continuous integration environments in order to triangulate the live collected data with information from testing the endpoints with a constant load and introduce a mechanism for performance prediction is added (Section \ref{sec:testing})
  \item A more extended discussion about lessons learned while developing the tool including measurements on the overhead incurred by the presented tool \ml{(... if we make it... )}
\end{itemize}

@@

  \Fref{fig:zeeguuarch} presents a high-level view of the case study ecosystem.  The core API is implemented with Flask and Python and provides three types of functionality, respectively: contextual translations, personalized exercise suggestions, and article recommendations. In total, the API provides a bit under 50 endpoints in total, out of which around a dozen are very frequently used. The development of the core API itself is a research project. 
  % One of the authors of this paper is part of the developer team that maintains and evolves the \zee API.

@@

  At the time of writing, the ecosystem consists of a reader web application, a web-based exercises platform, and a smartwatch application, which are used by more than \activeUserCount active users on a regular basis. The users come from several Dutch highschools and a language center associated with a Dutch University. Several users are using it on their own, outside of any formal educational context. The learners use \zee to study a variety of foreign languages including German, Dutch, and French. They have also diverse native languages, including: Chinese, Dutch, and English. The system can scale to many languages by delegating the machine translation tasks to third party APIs.

\subsection{Study characteristics}

@@

  All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API. The figures are interactively offering basic data exploration capabilities: filter, zoom, and details on demand~\cite{Shne99a}. The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em icsme}}.

@@

 Flask is built on top of WSGI\footnote{\url{http://wsgi.org}} --- The Web Server Gateway Interface --- a calling convention for web servers to forward requests to web applications or frameworks written in Python. A Python web application based on WSGI has to have one central callable object that implements the actual application. In Flask this is an instance of the Flask class. Each Flask application has to create an instance of this class itself and pass it the name of the module. The following code snippet presents a simple ``Hello World'' web service written in Python with the help of the Flask framework.

@@

In the following we discuss the capabilities of the \tool in more depth.
